---
title: "Causal Analysis"
author: "Alex Mansourati"
date: "04/05/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(boot)
library(tidyverse)
library(mgcv)
```

```{r}
# via http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```


```{r}
df = read.csv('output/nba_timeouts_analysis.csv')
df = df %>% select(-time_since_last_break)
head(df)
```

```{r}
df %>% group_by(treatment) %>% tally()
```



```{r}
df %>% ggplot(aes(x=outcome)) + geom_density()
```

# Assessment of Balance in Covariates between Treated and Controls
```{r}
library(gtable)
p = df %>%
  mutate(treatment=as.factor(treatment)) %>% 
  ggplot(aes(fill=treatment, alpha=0.2)) +
  ylab("")

p1 = p + geom_density(aes(x=scoring_run)) +theme(legend.position="none")
p2 = p + geom_density(aes(x=point_diff)) +theme(legend.position="none")
p3 = p + geom_density(aes(x=)) +theme(legend.position="none")
p4 = p + geom_density(aes(x=coach_exp)) 

multiplot(p1, p2, p3, cols=2)
```

The distribution look mostly the same, except for `scoring_run`. The scoring run covariate as seen in Figure XYZ, is a similarly shaped normal distribution (same SD) but translated slightly. We can handle this adjustment in our model. 

```{r}
a = df %>% 
  pivot_longer(-c(treatment, outcome)) %>% 
  group_by(treatment, name) %>% 
  summarise(mean = mean(value), sd = sd(value)) %>% 
  arrange(name, treatment) %>% 
  pivot_wider(names_from='treatment', values_from=c('mean', 'sd'))
```

```{r}
b = df %>% 
  pivot_longer(-c(treatment, outcome, coach_exp)) %>% 
  group_by(treatment, name) %>% 
  summarise(opp_lower_quant = quantile(value, 0.025),
            opp_upper_quant = quantile(value, 0.975)) %>% 
  mutate(opposite_treatment = case_when(treatment==0~1, TRUE~0)) %>%
  inner_join(df %>% pivot_longer(-c(treatment, outcome)), by=c("opposite_treatment"="treatment", "name")) %>%
  # inner_join(df %>% pivot_longer(-c(treatment, outcome)), by=c("treatment"="treatment", "name")) %>% 
  group_by(treatment, name) %>% 
  summarise(pi = ecdf(value)(first(opp_lower_quant)) + (1-ecdf(value)(first(opp_upper_quant)))) %>% 
  pivot_wider(names_from=c('treatment'), values_from=c('pi'), names_prefix='pi_0.05_')
```

```{r}
a %>% inner_join(b, by='name')
```




# Propensity Score Model using GAM

Following Shalizi's advivce on building a well-specified statistical model. 

```{r}
df
```


```{r}
lin_fit = glm(formula = treatment~scoring_run + coach_exp + point_diff,
   data=df, 
   family=binomial())

print(summary(lin_fit))
```

R^2: 
```{r}
with(summary(lin_fit), 1 - deviance/null.deviance)
```

```{r}
prop_score_df = tibble(predicted = predict(lin_fit, type='response'), 
                       actual = as.factor(df$treatment))
prop_score_df %>%
  ggplot(aes(x=predicted)) + 
  geom_histogram(color='white', bins=20) + 
  facet_wrap(~actual) +
  theme_bw()
```




# GAM
```{r}
library(mgcv)
np_fit = gam(treatment~s(scoring_run) + s(coach_exp) + s(point_diff),
             data=df, family=binomial())

summary(np_fit)
```

```{r}
plot(np_fit)
```


```{r}
prop_score_df = tibble(predicted = predict(np_fit, type='response'), 
                       actual = as.factor(df$treatment))
prop_score_df %>%
  ggplot(aes(x=predicted)) + 
  geom_histogram(color='white', bins=20) + 
  facet_wrap(~actual) +
  theme_bw()
```



# What is a typical game's timeout cadence?

Let's take a look at a couple games
```{r}
timeout_df = read_csv('output/timeouts.csv')
timeout_df %>% 
  filter(game_id == 21800742) %>% 
  ggplot(aes(x=elapsed_seconds, y=1)) + geom_point()
```

# Naive

.... take from first report


# Design Phase

Matching via Propensity Score 

MatchIt matches on propensity score, not the linearized propensity score. 

```{r}
library("MatchIt")
m = matchit(treatment ~ scoring_run + coach_exp + point_diff,
                 data = df, method = "nearest")
```

```{r}
summary(m$model)
```

```{r}
plot(m, type="QQ")
```

```{r}
plot(m, type="hist")
```


Let's calculate the ATT using the matched dataset. 

```{r}
# Calculate ATT
# See Shalizi page 458 or Imbens-Rubin page 407
cum_diff = 0
cum_diff_cate_pos = 0
cum_diff_cate_neg = 0
pos_n = 0
neg_n = 0
for (i in 1:nrow(m$match.matrix)){
  treated_outcome = df %>% slice(i) %>% select(outcome) %>% pull()
  control_outcome = df %>% slice(as.integer(m$match.matrix[i])) %>% select(outcome) %>% pull()
  diff = treated_outcome - control_outcome
  
  scoring_run = df %>% slice(as.integer(m$match.matrix[i])) %>% select(scoring_run) %>% pull()
  if (scoring_run > 0){
    cum_diff_cate_pos = cum_diff_cate_pos + diff
    pos_n = pos_n + 1
  }
  if (scoring_run < 0){
    cum_diff_cate_neg = cum_diff_cate_neg + diff
    neg_n = neg_n + 1
  }
  
  cum_diff = cum_diff + diff
}

att = cum_diff / nrow(m$match.matrix)
catt_pos = cum_diff_cate_pos / pos_n
catt_neg = cum_diff_cate_neg / neg_n

# Calculate variance of ATT 
# See Imbens-Rubin page 407
cum_var_term = 0
cum_var_term_cate_pos = 0
cum_var_term_cate_neg = 0
for (i in 1: nrow(m$match.matrix)){
  treated_outcome = df %>% slice(i) %>% select(outcome) %>% pull()
  control_outcome = df %>% slice(as.integer(m$match.matrix[i])) %>% select(outcome) %>% pull()
  diff = treated_outcome - control_outcome
  scoring_run = df %>% slice(as.integer(m$match.matrix[i])) %>% select(scoring_run) %>% pull()
  
  if (scoring_run > 0){
    pos_var_term = (diff - catt_pos)^2 / 2
    cum_var_term_cate_pos = cum_var_term_cate_pos + pos_var_term
  }
  if (scoring_run < 0){
    neg_var_term = (diff - catt_neg)^2 / 2
    cum_var_term_cate_neg = cum_var_term_cate_neg + neg_var_term
  }
  
  var_term = (diff - att)^2 / 2
  cum_var_term = cum_var_term + var_term
}

var_att = cum_var_term / nrow(m$match.matrix)
var_catt_pos = cum_var_term_cate_pos / pos_n
var_catt_neg = cum_var_term_cate_neg / neg_n
```

```{r}
print(paste(att, var_att^0.5))
print(paste('positive', catt_pos, var_catt_pos^0.5))
print(paste('negative', catt_neg, var_catt_neg^0.5))
```

Uses a conservative estimate of propensity score variance. Not seeing anything significant here. 


```{r}
treated = match.data(m, group = "treat")
control = match.data(m, group = "control")
n_treated = nrow(treated)
sum_treated = sum(treated$outcome)
n_control = nrow(control)
sum_control = sum(control$outcome)
(sum_treated / n_treated - sum_control / n_control)
```

```{r}
treated
```


# Modelling Phase
Do we have unit-level bias in our matching estimator? (Page 416 Imbens-Rubin)

The following code tests for unit-level bias. 

# ```{r}
# k = matrix(0, nrow=nrow(m$match.matrix), ncol=1)
# for (i in 1:nrow(m$match.matrix)){
#   treated_prop_score = logit(m$distance[i])
#   control_prop_score = logit(m$distance[as.integer(m$match.matrix[i])])
#   diff = treated_prop_score - control_prop_score
#   k[i] = diff
# }
# tibble(k) %>% ggplot(aes(x=k)) + geom_histogram()
# ```


We want to do outcome modelling to get an idea of variance (bootstrap for matching estimators is not a good technique).


```{r}
glinfit = lm(outcome ~ treatment + scoring_run + coach_exp + point_diff, data=match.data(m))
```

```{r}
summary(glinfit)
```



```{r}
library('hexbin')
k = tibble(fitted=glinfit$fitted.values, actual=match.data(m)$outcome, residuals=glinfit$residuals) 
k = k %>% sample_frac(0.05)

k %>% 
  ggplot(aes(x=fitted, y=residuals)) + 
  # stat_binhex()
  # geom_point()
  geom_jitter(height=0.25)
```


```{r}
df = match.data(m)
outcome_model = lm(outcome~treatment+scoring_run+coach_exp+point_diff, data=df)
df_all_treated = df
df_all_treated$treatment = 1
df_none_treated = df
df_none_treated$treatment = 0
df_standardization = rbind(df_all_treated, df_none_treated)
outcomes = predict(outcome_model, newdata=df_standardization)
treated_index_start = 0
treated_index_end = nrow(df_all_treated)
untreated_index_start = as.integer(nrow(df_all_treated) + 1)
untreated_index_end = nrow(df_standardization)
standardization_ACE = (sum(outcomes[treated_index_start:treated_index_end]) / nrow(df_all_treated)) - (sum(outcomes[untreated_index_start:untreated_index_end]) / nrow(df_none_treated))
```


```{r}
standardization_ACE
```

```{r}
fc <- function(df,i){
  df_boot <- df[i,]
  outcome_model = lm(outcome~treatment+scoring_run+coach_exp+point_diff, data=df_boot)
  df_all_treated = df_boot
  df_all_treated$treatment = 1
  df_none_treated = df_boot
  df_none_treated$treatment = 0
  df_standardization = rbind(df_all_treated, df_none_treated)
  outcomes = predict(outcome_model, newdata=df_standardization)
  treated_index_start = 0
  treated_index_end = nrow(df_all_treated)
  untreated_index_start = as.integer(nrow(df_all_treated) + 1)
  untreated_index_end = nrow(df_standardization)
  standardization_ACE = (sum(outcomes[treated_index_start:treated_index_end]) / nrow(df_all_treated)) - (sum(outcomes[untreated_index_start:untreated_index_end]) / nrow(df_none_treated))
  return(standardization_ACE)
}

boot_standardization_ACE <- boot(df, fc, R=500)
standardization_CI = toString(standardization_ACE + c(-1,1) * 1.96 * sd(boot_standardization_ACE$t))
```

```{r}
standardization_CI
```



# Positive CATT
```{r}
df_positive = df[df$scoring_run > 0,]

outcome_model = lm(outcome~treatment+scoring_run+coach_exp+point_diff, data=df_positive)
df_positive_all_treated = df_positive
df_positive_all_treated$treatment = 1
df_positive_none_treated = df_positive
df_positive_none_treated$treatment = 0
df_positive_standardization = rbind(df_positive_all_treated, df_positive_none_treated)
outcomes = predict(outcome_model, newdata=df_positive_standardization)
treated_index_start = 0
treated_index_end = nrow(df_positive_all_treated)
untreated_index_start = as.integer(nrow(df_positive_all_treated) + 1)
untreated_index_end = nrow(df_positive_standardization)
standardization_ACE = (sum(outcomes[treated_index_start:treated_index_end]) / nrow(df_positive_all_treated)) - (sum(outcomes[untreated_index_start:untreated_index_end]) / nrow(df_positive_none_treated))

```


```{r}
fc <- function(df_positive,i){
  df_positive_boot <- df_positive[i,]
  outcome_model = lm(outcome~treatment+scoring_run+coach_exp+point_diff, data=df_positive_boot)
  df_positive_all_treated = df_positive_boot
  df_positive_all_treated$treatment = 1
  df_positive_none_treated = df_positive_boot
  df_positive_none_treated$treatment = 0
  df_positive_standardization = rbind(df_positive_all_treated, df_positive_none_treated)
  outcomes = predict(outcome_model, newdata=df_positive_standardization)
  treated_index_start = 0
  treated_index_end = nrow(df_positive_all_treated)
  untreated_index_start = as.integer(nrow(df_positive_all_treated) + 1)
  untreated_index_end = nrow(df_positive_standardization)
  standardization_ACE = (sum(outcomes[treated_index_start:treated_index_end]) / nrow(df_positive_all_treated)) - (sum(outcomes[untreated_index_start:untreated_index_end]) / nrow(df_positive_none_treated))
  return(standardization_ACE)
}

boot_standardization_ACE <- boot(df_positive, fc, R=500)
standardization_CI = toString(standardization_ACE + c(-1,1) * 1.96 * sd(boot_standardization_ACE$t))
```

```{r}
standardization_ACE
```


```{r}
standardization_CI
```


#Negative CATT
```{r}
df_negative = df[df$scoring_run < 0,]

outcome_model = lm(outcome~treatment+scoring_run+coach_exp+point_diff, data=df_negative)
df_negative_all_treated = df_negative
df_negative_all_treated$treatment = 1
df_negative_none_treated = df_negative
df_negative_none_treated$treatment = 0
df_negative_standardization = rbind(df_negative_all_treated, df_negative_none_treated)
outcomes = predict(outcome_model, newdata=df_negative_standardization)
treated_index_start = 0
treated_index_end = nrow(df_negative_all_treated)
untreated_index_start = as.integer(nrow(df_negative_all_treated) + 1)
untreated_index_end = nrow(df_negative_standardization)
standardization_ACE = (sum(outcomes[treated_index_start:treated_index_end]) / nrow(df_negative_all_treated)) - (sum(outcomes[untreated_index_start:untreated_index_end]) / nrow(df_negative_none_treated))

```


```{r}
fc <- function(df_negative,i){
  df_negative_boot <- df_negative[i,]
  outcome_model = lm(outcome~treatment+scoring_run+coach_exp+point_diff, data=df_negative_boot)
  df_negative_all_treated = df_negative_boot
  df_negative_all_treated$treatment = 1
  df_negative_none_treated = df_negative_boot
  df_negative_none_treated$treatment = 0
  df_negative_standardization = rbind(df_negative_all_treated, df_negative_none_treated)
  outcomes = predict(outcome_model, newdata=df_negative_standardization)
  treated_index_start = 0
  treated_index_end = nrow(df_negative_all_treated)
  untreated_index_start = as.integer(nrow(df_negative_all_treated) + 1)
  untreated_index_end = nrow(df_negative_standardization)
  standardization_ACE = (sum(outcomes[treated_index_start:treated_index_end]) / nrow(df_negative_all_treated)) - (sum(outcomes[untreated_index_start:untreated_index_end]) / nrow(df_negative_none_treated))
  return(standardization_ACE)
}

boot_standardization_ACE <- boot(df_negative, fc, R=500)
standardization_CI = toString(standardization_ACE + c(-1,1) * 1.96 * sd(boot_standardization_ACE$t))
```

```{r}
standardization_ACE
```


```{r}
standardization_CI
```


